---
title: "The DMRscaler user's guide"
author: "Leroy Bondhus"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The DMRscaler user's guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Summary
DMRscalar


## Example Dataset Setup
Here we'll use a relatively small dataset from GEO, 

```{r}
library(GEOquery)
library(minfi)

## NOTE: 170.2 MB 
gse_id <- "GSE149960"
gse <- getGEO(gse_id, GSEMatrix = TRUE)
phen <- gse[[1]]@phenoData@data

getGEOSuppFiles(gse_id)
idats_dir<-paste(g"GSE149960/idat"
untar("GSE149960/GSE149960_RAW.tar", exdir = "GSE149960/idat")
head(list.files("GSE149960/idat", pattern = "idat"))

idatFiles <- list.files("GSE149960/idat", pattern = "idat.gz$", full = TRUE)
sapply(idatFiles, gunzip, overwrite = TRUE)

###  Reading of idat files done with minfi library ###
idats_dir<-"GSE149960/idat"
idats_files<-list.files(path=idats_dir, pattern = "*.idat")
targets <- unlist(strsplit(list.files("GSE149960/idat", pattern = "idat"), split="_[RG][er][dn]"))
targets <- targets[which(targets != ".idat.gz")]
RGSet <- read.metharray.exp(base = idats_dir)
GRset.funnorm <- preprocessFunnorm(RGSet);rm(RGSet)
snps <- getSnpInfo(object = GRset.funnorm)
GRset.funnorm <- dropLociWithSnps(GRset.funnorm, snps=c("SBE", "CpG"), maf=0);rm(snps)
rm(idats_dir, idats_files, targets)


#### GSE dat appears to be same as minfi processed data
#hist(B[rownames(dat),]-dat[])


```







Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
