---
title: "The DMRscaler user's guide"
author: "Leroy Bondhus"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The DMRscaler user's guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
                      warning=FALSE)
```
## Summary
DMRscaler is a method designed to identify features of differential methylation between case and control conditions across a wide range of genomic scale. In the first part of the vignette we step through downloading a publicly available dataset, running the statistical test for difference at individual CG site and a permutation test for estimating enrichment of differentially methylated CGs within a window. After this we run DMRscaler to call DMRs at each layer of window sizes generating a list of results. We then look at the data to visualize how DMRscaler has called and represents these DMR features.       


## install DMRscaler 
To install pull the most recent version from github and place in working directory 
```{r, results='hide', warning=FALSE, , message=FALSE}
library("devtools")
library("roxygen2")
document()
install("../DMRscaler")
```

## Example Dataset Setup
We will use data from GSE149960 from (REFERENCE) with DNA methylation from fibroblasts from progeria patients and controls measured on the Illumina methylation EPIC array.  
```{r get-dataset-dependencies, message=FALSE}
library("GEOquery")
```

```{r get-dataset, results='hide'}
## get sample phenotype data table 
gse <- getGEO("GSE149960", GSEMatrix = TRUE)
phen <- gse$GSE149960_series_matrix.txt.gz@phenoData@data
rm(gse)

## get methylation data as idat files (NOTE: this saves files locally in working directory, unpacked size is 411 Mb)
## unpacked size is 411 Mb  
getGEOSuppFiles("GSE149960")
untar("GSE149960/GSE149960_RAW.tar", exdir = "GSE149960/idat")
file.remove("GSE149960/GSE149960_RAW.tar")

idat_files <- list.files("GSE149960/idat", pattern = "idat.gz$", full = TRUE)
sapply(idat_files, gunzip, overwrite = TRUE); rm(idat_files)
```



##  Preprocessing
Preprocessing of idat files is done with minfi here.
```{r preprocessing-dependencies, message=FALSE}
library("minfi")
```

```{r preprocessing}
###  Reading of idat files done with minfi library ###
idats_dir<-"GSE149960/idat"
RGSet <- read.metharray.exp(base = idats_dir)
GRset.funnorm <- preprocessFunnorm(RGSet);rm(RGSet)
snps <- getSnpInfo(object = GRset.funnorm)
GRset.funnorm <- dropLociWithSnps(GRset.funnorm, snps=c("SBE", "CpG"), maf=0);rm(snps)
rm(idats_dir)
```


## Set up for DMRscaler
```{r data-setup-dependencies, message=FALSE}
library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
```

```{r data_setup}
controls <- grep("control",phen$title)
cases <- grep("hgps", phen$title)
locs <- getLocations(GRset.funnorm)
locs <- data.frame("names"=locs@ranges@NAMES, "pos"=locs@ranges@start, "chr" = rep(locs@seqnames@values, locs@seqnames@lengths))
B <- getBeta(GRset.funnorm)
```


## Run statistical tests and permutations to feed into DMRscaler method
We use the non-parametric wilcox test for individual CG significance calculations and permutation of these significance values for estimating window level significance. 
ToDo: The permutations involved are time consuming (~3-5 min per permutation, should run with at least 10), so the permutations should be run and the results included with the package as example data.   
```{r DMRscaler-set-up-dependencies, message=FALSE}
library("doParallel")
library("rlang")
library("dplyr")
library("foreach")
```

```{r DMRscaler-set-up}
registerDoParallel(detectCores())

num_perm <- 10
clt_reps <- 1e4

rim <- DMRscaler::generate_rand_index_matrix(num_controls = length(controls),
                                             num_cases = length(cases),
                                             num_permutations = num_perm)
mrp <- DMRscaler::run_MWW_rand_permutation(index_matrix = rim, 
                                           Beta = B,
                                           num_permutations = num_perm)
mrp <- -log10(mrp)
mwr <- DMRscaler::run_MWW(control_indices = controls ,
                          case_indices = cases ,
                          Beta = B)
mwr <- -log10(mwr)
fdt <- DMRscaler::write_FDR_table(real_table = mwr,
                                  rand_table = mrp)

fdrscaler <- DMRscaler::get_FDR_scalar(MWW_FDR_table = fdt,
                                       MWW_FDR_threshold = 0.1)
if(is.na(fdrscaler)){fdrscaler <- 1}

## This is where significance level of individual windows is estimated by permutation 
cltable <- DMRscaler::write_CLT_lookup_table(num_reps = clt_reps ,
                                             data_to_sample = mwr$p_val,
                                             FDR_scaler = fdrscaler,
                                             clt_numCGs = c(2, 5, 10, 25, 50))
locs$chr <- as.factor(locs$chr)
locs$scoring_values <- mwr$p_val
```


## Run DMRscaler to call DMRs
Uses significance values associated with each CG location and the values from the statistical test and permutations along with windows of adjacent measured CGs defined by the layer_sizes
ToDo: This can take   
```{r}
DMRscaler_result <- DMRscaler::dmrscaler(locs = locs, fdrscaler = fdrscaler, cltable=cltable, layer_sizes = c(4,8,16,32,64))
```

## Choose an example region to look at
```{r}
example_region <- data.frame(chr="chr7",start=155000000, stop=159000000, stringsAsFactors = FALSE)
```

```{r hilbert-curve-dependencies, message=FALSE}
library(circlize)
library(HilbertCurve)
```

```{r, fig.width = 4, fig.height=4}
col_fun = colorRamp2(c(0,fdrscaler*0.99,fdrscaler,max(locs$scoring_values)), c("grey30", "grey60", "red", "red"))
level = 6
## 4^level - 1 = # segments

hc_points <- locs[which(locs$chr== example_region$chr ),]
hc_max <- nrow(hc_points)
hc_col <- col_fun(hc_points$scoring_values)
hc_size <- hc_points$scoring_values / max(hc_points$scoring_values)

#  hc_size <- hc_points$scoring_values / fdrscaler
hc <- HilbertCurve(s=1,e=hc_max, level = level, reference = F, title = example_region$chr)
hc_points(hc, x1 = 1:nrow(hc_points), np = NULL, pch=15, size = unit(hc_size*2, "mm"),gp = gpar(col = hc_col, fill = hc_col))
hc_polygon(hc, x1 = min(which(hc_points$pos >= example_region$start)), x2 = max(which(hc_points$pos <= example_region$stop)) )

### now looking only at the specified region
level = 4
hc_points <- locs[which(locs$chr== example_region$chr & locs$pos >= example_region$start & locs$pos <= example_region$stop ),]
hc_max <- nrow(hc_points)
hc_col <- col_fun(hc_points$scoring_values)
hc_size <- hc_points$scoring_values / max(hc_points$scoring_values)

#  hc_size <- hc_points$scoring_values / fdrscaler
hc <- HilbertCurve(s=1,e=hc_max, level = level, reference = F, title = example_region$chr)
hc_points(hc, x1 = 1:nrow(hc_points), np = NULL, pch=15, size = unit(hc_size*4, "mm"),gp = gpar(col = hc_col, fill = hc_col))
```

## Explore some features of the data
DMRscaler defines differential methylation features iteratively, expanding the size of the window for aggregating on at each step
this procedure allows a hierarchical structure to describe a region enriched in differntially methylated CpGs. We look at our example region here 
```{r dmr-structure-visual-dependencies, results='hide'}
library(networkD3)
```

```{r dmr-structure-visual, fig.height=6, fig.width = 7}
dmr_tree <- example_generate_dmr_tree(DMRscaler_result = DMRscaler_result, layer=5, chr=example_region$chr, start = example_region$start, stop = example_region$stop)
diagonalNetwork(List = dmr_tree, fontSize = 12, fontFamily = "bold", nodeStroke = "black", linkColour = "black", opacity = 1 )

```

Next we can look at the region using the Gviz package
```{r genenomic-range-visual-dependencies}
library(Gviz)
library(biomaRt)
```

```{r genomic-range-visual, fig.width = 7, fig.height=5}
## organize data used in tracks
group <- character(length=ncol(B)) 
group[cases] <- "case"
group[controls] <- "control"
delta_mean_B <- rowMeans(B[,cases]) - rowMeans(B[,controls]) 
which <- which(locs$chr==example_region$chr & locs$pos >= example_region$start & locs$pos <= example_region$stop)
Bsub <- B[which, ]
gr <- GRanges(seqnames = example_region$chr, ranges = IRanges(start = locs[which,"pos"], width = 1), mcols = Bsub )

## set up ideogram track
itrack<-IdeogramTrack(genome="hg19", chromosome = example_region$chr)
## set up genome axis track
gtrack<-GenomeAxisTrack()
## set up significance track
sig_track <- DataTrack(range=gr, data = locs$scoring_values[which], type=c("p"))
## set up beta value track
beta_track <- DataTrack(range = gr, data = t(Bsub), groups=group,name="Beta", type=c("a","g","confint"))
## set up diff beta track
diff_beta_track <- DataTrack(range = gr, data = delta_mean_B[which], type=c("a","g"))

## set up gene model track ### 
genesub <- example_gene_anno_df(chr=example_region$chr, start=example_region$start, stop=example_region$stop)
if(!all(is.na(genesub))){
  grtrack <- GeneRegionTrack(genesub, chromosome = example_region$chr, start = example_region$start, stop=example_region$stop, transcriptAnnotation="symbol" ,collapseTranscripts = "meta")
}else{grtrack<-GeneRegionTrack()}

## set up DMRscaler results layer tracks
layer_track <- list()
for(j in 1:length(DMRscaler_result)){
  layer_track[[j]] <- AnnotationTrack(start = DMRscaler_result[[j]]$start_pos, 
                                      end =  DMRscaler_result[[j]]$stop_pos,
                                      chromosome = DMRscaler_result[[j]]$chr,
                                      strand = "*", genome = "hg19", name=paste("layer", j, sep = "_"))
  displayPars(layer_track[[j]]) <- list(cex.legend=0.7, cex.axis=0.7, cex.title=0.7, rotation.title=0, stackHeight = 1, shape="box")
}


plotTracks(c(list(itrack, gtrack, sig_track, beta_track, diff_beta_track, grtrack), layer_track
                ), from = example_region$start-1, to=example_region$stop+1 ) 
```


```{r}
sessionInfo()

```
