---
title: "The DMRscaler user's guide"
author: "Leroy Bondhus"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The DMRscaler user's guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Summary
DMRscalar


## Example Dataset Setup
We will use data from GSE149960 from (REFERENCE) with DNA methylation from fibroblasts from progeria patients and controls measured on the Illumina methylation EPIC array.  
### Note: This vignette imports files from the Gene Expression Omnibus. (TODO: set cleanup option to remove files)
```{r get_data }
library(GEOquery)

## Change this line to change base directory where files will be stored
base_dir <- getwd()

## NOTE: 170.2 MB 
gse_id <- "GSE149960"
gse <- getGEO(gse_id, GSEMatrix = TRUE)
phen <- gse[[1]]@phenoData@data; rm(gse)
## NOTE: 214.6 MB, saves files in base_dir/GSE149960
geo_file <- getGEOSuppFiles(gse_id, baseDir = base_dir)
geo_file <- rownames(geo_file)

idat_dir <- paste(base_dir, gse_id, "idat", sep = "/")
untar(geo_file, exdir = idat_dir)
file.remove(geo_file) ## delete the *.tar file

idat_files <- list.files(idat_dir, full = TRUE)
sapply(idat_files, gunzip, overwrite = TRUE)  ## unzip the files

```


###  Preprocessing idat files with minfi 
```{r preprocessing_data}
library(minfi)
RGSet <- read.metharray.exp(base = idat_dir)
GRset.funnorm <- preprocessFunnorm(RGSet);rm(RGSet) ## This can take a few minutes
snps <- getSnpInfo(object = GRset.funnorm)
GRset.funnorm <- dropLociWithSnps(GRset.funnorm, snps=c("SBE", "CpG"), maf=0);rm(snps)
#rm(idats_dir, idats_files, targets)

```


```{r data_setup}
controls <- grep("control",phen$title)
cases <- grep("hgps", phen$title)

locs <- getLocations(GRset.funnorm)
locs <- data.frame("names"=locs@ranges@NAMES,
                   "pos"=locs@ranges@start,
                   "chr" = rep(locs@seqnames@values, locs@seqnames@lengths))
B <- getBeta(GRset.funnorm)

```


### install dmrscaler (BIOCONDUCTOR SUBMISSION IN PROGRESS)
```{r}



```




Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
